\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-23 papers/jss/article.Rnw
Liberal sprinkling of \proglang{} anywhere we mention R, C++, Python, or Java.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@825 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{
Modern data collection and analysis pipelines often involve
a sophisticated mix of applications written in general purpose and
specialized programming languages.  Protocol Buffers are a popular
method of serializing structured data between applications---while remaining
independent of programming languages or operating system.  The
\CRANpkg{RProtoBuf} package provides a complete interface between this
library and the R environment for statistical computing.
%TODO(ms) keep it less than 150 words.
% Maybe add Jeroen's sentence:
% JO: added this sentence to the conclustion, but could use it in abstract as well.
% They offer a unique combination of features, performance, and maturity that seems
% particulary well suited for data-driven applications and numerical
% computing.
}
\Keywords{\proglang{R}, \pkg{Rcpp}, protocol buffers, serialization, cross-platform}
\Plainkeywords{r, Rcpp, protocol buffers, serialization, cross-platform} %% without formatting
%% at least one keyword must be supplied

@@

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character separated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as separator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexity at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.
%
A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is inefficient and
leads to loss of type (double versus integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R
objects \citep{rjson,RJSONIO,jsonlite}.

@@

This paper describes an R interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \CRANpkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

%This article describes the basics of Google's Protocol Buffers through
%an easy to use R package, \CRANpkg{RProtoBuf}.  After describing the
%basics of protocol buffers and \CRANpkg{RProtoBuf}, we illustrate
%several common use cases for protocol buffers in data analysis.

@@

Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication workflow with Protocol Buffers and an interactive R session.
Common use cases include populating a request remote-procedure call (RPC)
Protocol Buffer in R that is then serialized and sent over the network to a
remote server.  The server would then deserialize the message, act on the
request, and respond with a new Protocol Buffer over the network. 
The key difference to, say, a request to an Rserve instance is that
the remote server may be implemented in any language, with no
dependence on R.

While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is flexible and simple to use.  The schema for structured
protocol buffer data is defined in \texttt{.proto} files which may
contain one or more message types.  Each message type has one or more
fields.  A field is specified with a unique number, a name, a value
type, and a field rule specifying whether the field is optional,
required, or repeated.  The supported value types are numbers,
enumerations, booleans, strings, raw bytes, or other nested message
types.  The \texttt{.proto} file syntax for defining the structure of protocol
buffer data is described comprehensively on Google Code\footnote{See 
\url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.}.
Table~\ref{tab:proto} shows an example \texttt{.proto} file which
defines the \texttt{tutorial.Person} type.  The R code in the right
column shows an example of creating a new message of this type and
populating its fields.

%% TODO(de) Can we make this not break the width of the page?
\noindent
\begin{table}
\begin{tabular}{p{.40\textwidth}p{0.55\textwidth}}
\toprule
Schema : \texttt{addressbook.proto} & Example R Session\\
\cmidrule{1-2}
\begin{minipage}{.40\textwidth}
\vspace{2mm}
\begin{example}
package tutorial;
message Person {
 required string name = 1;
 required int32 id = 2;
 optional string email = 3;
 enum PhoneType {
   MOBILE = 0; HOME = 1;
   WORK = 2;
 }
 message PhoneNumber {
   required string number = 1;
   optional PhoneType type = 2;
 }
 repeated PhoneNumber phone = 4;
}
\end{example}
\vspace{2mm}
\end{minipage} & \begin{minipage}{.55\textwidth}
<<echo=TRUE>>=
library(RProtoBuf)
p <- new(tutorial.Person,id=1,name="Dirk")
p$name
p$name <- "Murray"
cat(as.character(p))
serialize(p, NULL)
class(p)
@
\end{minipage} \\
\bottomrule
\end{tabular}
\caption{The schema representation from a \texttt{.proto} file for the
  \texttt{tutorial.Person} class (left) and simple R code for creating
  an object of this class and accessing its fields (right).}
\label{tab:proto}
\end{table}

@@

For added speed and efficiency, the C++, Java, and Python bindings to
Protocol Buffers are used with a compiler that translates a Protocol
Buffer schema description file (ending in \texttt{.proto}) into
language-specific classes that can be used to create, read, write and
manipulate Protocol Buffer messages.  The R interface, in contrast,
uses a reflection-based API that is particularly well-suited for
interactive data analysis.  
All messages in R have a single class
structure, but different accessor methods are created at runtime based
on the named fields of the specified message type, as described in the
next section.

@@

This section describes how to use the R API to create and manipulate
protocol buffer messages in R, and how to read and write the
binary representation of the message (often called the \emph{payload}) to files and arbitrary binary
R connections.
The two fundamental building blocks of Protocol Buffers are \emph{Messages}
and \emph{Descriptors}.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.

@@

\subsection{Importing Message Descriptors from .proto files}

@@

To create or parse a Protocol Buffer Message, one must first read in 
the message type specification from a \texttt{.proto} file. The 
\texttt{.proto} files are imported using the \code{readProtoFiles}
function, which can either import a single file, all files in a directory,
or every \texttt{.proto} file provided by a particular R package.

After importing proto files, the corresponding message descriptors are
available from the \texttt{RProtoBuf:DescriptorPool} environment in 
the R search path.  This environment is implemented with the user
defined tables framework from the \pkg{RObjectTables} package
available from the OmegaHat project \citep{RObjectTables}.  Instead of
being associated with a static hash table, this environment
dynamically queries the in-memory database of loaded descriptors
during normal variable lookup.

@@

% The old section 7 in entirety:
%The \texttt{RProtoBuf} package uses the user defined tables framework
%that is defined as part of the \texttt{RObjectTables} package available
%from the OmegaHat project \citep{RObjectTables}.
%
%The feature allows \texttt{RProtoBuf} to install the
%special environment \emph{RProtoBuf:DescriptorPool} in the R search path.
%The environment is special in that, instead of being associated with a
%static hash table, it is dynamically queried by R as part of R's usual
%variable lookup. In other words, it means that when the R interpreter
%looks for a binding to a symbol (foo) in its search path,
%it asks to our package if it knows the binding "foo", this is then
%implemented by the \texttt{RProtoBuf} package by calling an internal
%method of the \texttt{protobuf} C++ library.

%\subsection{Importing proto files}
%In contrast to the other languages (Java, C++, Python) that are officially
%supported by Google, the implementation used by the \texttt{RProtoBuf}
%package does not rely on the \texttt{protoc} compiler (with the exception of
%the two functions discussed in the previous section). This means that no
%initial step of statically compiling the proto file into C++ code that is
%then accessed by R code is necessary. Instead, \texttt{proto} files are
%parsed and processed \textsl{at runtime} by the protobuf C++ library---which
%is much more appropriate for a dynamic language.

@@

However, as opposed to R lists, no partial matching is performed
and the name must be given entirely.
The \verb|[[| operator can also be used to query and set fields
of a messages, supplying either their name or their tag number :

@@

Protocol Buffers include a 64-bit integer type, but R lacks native
64-bit integer support.  A workaround is available and described in
Section~\ref{sec:int64} for working with large integer values.

@@

However, the main focus of Protocol Buffer messages is
efficiency. Therefore, messages are transported as a sequence
of bytes. The \texttt{serialize} method is implemented for
Protocol Buffer messages to serialize a message into a sequence of
bytes that represents the message.
%(raw vector in R speech) that represents the message.

@@

The \texttt{input} argument of \texttt{read} can also be a binary
readable R connection, such as a binary file connection:

@@

The \CRANpkg{RProtoBuf} package uses the S4 system to store
information about descriptors and messages.  Using the S4 system
allows the package to dispatch methods that are not
generic in the S3 sense, such as \texttt{new} and
\texttt{serialize}.
Table~\ref{class-summary-table} lists the six
primary Message and Descriptor classes in RProtoBuf.  Each R object
contains an external pointer to an object managed by the
\texttt{protobuf} C++ library, and the R objects make calls into more
than 100 C++ functions that provide the
glue code between the R language classes and the underlying C++
classes.

@@

The \CRANpkg{Rcpp} package
\citep{eddelbuettel2011rcpp,eddelbuettel2013seamless} is used to 
facilitate this integration of the R and C++ code for these objects.
Each method is wrapped individually which alllows us to add user
friendly custom error handling, type coercion, and performance
improvements at the cost of a more verbose implementation.
The RProtoBuf package in many ways motivated
the development of Rcpp Modules \citep{eddelbuettel2013exposing},
which provide a more concise way of wrapping C++ functions and classes
in a single entity.

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Slot} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{pointer} & External pointer to the \texttt{Message} object of the C++ protobuf library. Documentation for the
\texttt{Message} class is available from the Protocol Buffer project page. \\
%(\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.message.html#Message}) \\
\texttt{type} & Fully qualified name of the message. For example a \texttt{Person} message
has its \texttt{type} slot set to \texttt{tutorial.Person} \\[.3cm]
\textbf{Method} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{has} & Indicates if a message has a given field.   \\
\texttt{clone} & Creates a clone of the message \\
\texttt{isInitialized} & Indicates if a message has all its required fields set\\
\texttt{serialize} & serialize a message to a file, binary connection, or raw vector\\
\texttt{clear} & Clear one or several fields of a message, or the entire message\\
\texttt{size} & The number of elements in a message field\\
\texttt{bytesize} & The number of bytes the message would take once serialized\\[3mm]
%
\texttt{swap} & swap elements of a repeated field of a message\\
\texttt{set} & set elements of a repeated field\\
\texttt{fetch} & fetch elements of a repeated field\\
\texttt{setExtension} & set an extension of a message\\
\texttt{getExtension} & get the value of an extension of a message\\
\texttt{add} & add elements to a repeated field \\[3mm]
%
\texttt{str} & the R structure of the message\\
\texttt{as.character} & character representation of a message\\
\texttt{toString} & character representation of a message (same as \texttt{as.character}) \\
\texttt{as.list} & converts message to a named R list\\
\texttt{update} & updates several fields of a message at once\\
\texttt{descriptor} & get the descriptor of the message type of this message\\
\texttt{fileDescriptor} & get the file descriptor of this message's descriptor\\
\hline
\end{tabular}
\end{small}
\caption{\label{Message-methods-table}Description of slots and methods for the \texttt{Message} S4 class}
\end{table}

@@

Descriptors describe the type of a Message.  This includes what fields
a message contains and what the types of those fields are.  Message
descriptors are represented in R with the \emph{Descriptor} S4
class. The class contains the slots \texttt{pointer} and
\texttt{type}.  Similarly to messages, the \verb|$| operator can be
used to retrieve descriptors that are contained in the descriptor, or
invoke pseudo-methods.

@@

\begin{table}[tbp]
\centering
\begin{small}
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Slot} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{pointer} & External pointer to the \texttt{Descriptor} object of the C++ proto library. Documentation for the
\texttt{Descriptor} class is available from the Protocol Buffer project page.\\
%\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\texttt{type} & Fully qualified path of the message type. \\[.3cm]
%
\textbf{Method} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{as.list} & return a named
list of the field, enum, and nested descriptors included in this descriptor.\\
\texttt{asMessage} & return DescriptorProto message. \\
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested 
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\bottomrule
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of slots and methods for the \texttt{Descriptor} S4 class}
\end{table}

@@

The class \emph{FieldDescriptor} represents field
descriptors in R. This is a wrapper S4 class around the
\texttt{google::protobuf::FieldDescriptor} C++ class.
Table~\ref{fielddescriptor-methods-table} describes the methods
defined for the \texttt{FieldDescriptor} class.

\begin{table}[tbp]
\centering
\begin{small}
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Slot} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{pointer} & External pointer to the \texttt{FieldDescriptor} C++ variable \\
\texttt{name} & Simple name of the field \\
\texttt{full\_name} & Fully qualified name of the field \\
\texttt{type} & Name of the message type where the field is declared \\[.3cm]
%
\textbf{Method} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{as.character} & Character representation of a descriptor\\
\texttt{toString} & Character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & Return FieldDescriptorProto message. \\
\texttt{name} & Return the name of the field descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{is\_extension} & Return TRUE if this field is an extension.\\
\texttt{number} & Gets the declared tag number of the field.\\
\texttt{type} & Gets the type of the field.\\
\texttt{cpp\_type} & Gets the C++ type of the field.\\
\texttt{label} & Gets the label of a field (optional, required, or repeated).\\
\texttt{is\_repeated} & Return TRUE if this field is repeated.\\
\texttt{is\_required} & Return TRUE if this field is required.\\
\texttt{is\_optional} & Return TRUE if this field is optional.\\
\texttt{has\_default\_value} & Return TRUE if this field has a default value.\\
\texttt{default\_value} & Return the default value.\\
\texttt{message\_type} & Return the message type if this is a message type field.\\
\texttt{enum\_type} & Return the enum type if this is an enum type field.\\
\bottomrule
\end{tabular}
\end{small}
\caption{\label{fielddescriptor-methods-table}Description of slots and
  methods for the \texttt{FieldDescriptor} S4 class}
\end{table}

@@

If we try to set an int64 field in R to double values, we lose
precision:

@@

The protocol buffer can then be read into R and converted to a native
R histogram object for plotting:

>>>>

%% an abstract and keywords
\Abstract{
Modern data collection and analysis pipelines often involve
a sophisticated mix of applications written in general purpose and
specialized programming languages.  Protocol Buffers are a popular
method of serializing structured data between applications---while remaining
independent of programming languages or operating system.  The
\CRANpkg{RProtoBuf} package provides a complete interface between this
library and the \proglang{R} environment for statistical computing.
%TODO(ms) keep it less than 150 words.
% Maybe add Jeroen's sentence:
% JO: added this sentence to the conclustion, but could use it in abstract as well.
% They offer a unique combination of features, performance, and maturity that seems
% particulary well suited for data-driven applications and numerical
% computing.
}
\Keywords{\proglang{R}, \pkg{Rcpp}, protocol buffers, serialization, cross-platform}
\Plainkeywords{R, Rcpp, protocol buffers, serialization, cross-platform} %% without formatting
%% at least one keyword must be supplied

@@

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character separated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as separator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexity at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.
%
A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is inefficient and
leads to loss of type (double versus integer) and precision. Several \proglang{R} packages
implement functions to parse and generate \texttt{JSON} data from \proglang{R}
objects \citep{rjson,RJSONIO,jsonlite}.

@@

This paper describes an \proglang{R} interface to Protocol Buffers,
and is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive \proglang{R} interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between \proglang{R} and other languages.  Section~\ref{sec:evaluation} introduces a
general \proglang{R} language schema for serializing arbitrary \proglang{R} objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:mapreduce}
and \ref{sec:opencpu} provide real-world use cases of \CRANpkg{RProtoBuf}
in MapReduce and web service environments, respectively, before
Section~\ref{sec:summary} concludes.

%This article describes the basics of Google's Protocol Buffers through
%an easy to use \proglang{R} package, \CRANpkg{RProtoBuf}.  After describing the
%basics of protocol buffers and \CRANpkg{RProtoBuf}, we illustrate
%several common use cases for protocol buffers in data analysis.

@@

Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication workflow with Protocol Buffers and an interactive \proglang{R} session.
Common use cases include populating a request remote-procedure call (RPC)
Protocol Buffer in \proglang{R} that is then serialized and sent over the network to a
remote server.  The server would then deserialize the message, act on the
request, and respond with a new Protocol Buffer over the network. 
The key difference to, say, a request to an Rserve instance is that
the remote server may be implemented in any language, with no
dependence on R.

While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is flexible and simple to use.  The schema for structured
protocol buffer data is defined in \texttt{.proto} files which may
contain one or more message types.  Each message type has one or more
fields.  A field is specified with a unique number, a name, a value
type, and a field rule specifying whether the field is optional,
required, or repeated.  The supported value types are numbers,
enumerations, booleans, strings, raw bytes, or other nested message
types.  The \texttt{.proto} file syntax for defining the structure of protocol
buffer data is described comprehensively on Google Code\footnote{See 
\url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.}.
Table~\ref{tab:proto} shows an example \texttt{.proto} file which
defines the \texttt{tutorial.Person} type.  The \proglang{R} code in the right
column shows an example of creating a new message of this type and
populating its fields.

%% TODO(de) Can we make this not break the width of the page?
\noindent
\begin{table}
\begin{tabular}{p{.40\textwidth}p{0.55\textwidth}}
\toprule
Schema : \texttt{addressbook.proto} & Example \proglang{R} Session\\
\cmidrule{1-2}
\begin{minipage}{.40\textwidth}
\vspace{2mm}
\begin{example}
package tutorial;
message Person {
 required string name = 1;
 required int32 id = 2;
 optional string email = 3;
 enum PhoneType {
   MOBILE = 0; HOME = 1;
   WORK = 2;
 }
 message PhoneNumber {
   required string number = 1;
   optional PhoneType type = 2;
 }
 repeated PhoneNumber phone = 4;
}
\end{example}
\vspace{2mm}
\end{minipage} & \begin{minipage}{.55\textwidth}
<<echo=TRUE>>=
library(RProtoBuf)
p <- new(tutorial.Person,id=1,name="Dirk")
p$name
p$name <- "Murray"
cat(as.character(p))
serialize(p, NULL)
class(p)
@
\end{minipage} \\
\bottomrule
\end{tabular}
\caption{The schema representation from a \texttt{.proto} file for the
  \texttt{tutorial.Person} class (left) and simple \proglang{R} code for creating
  an object of this class and accessing its fields (right).}
\label{tab:proto}
\end{table}

@@

For added speed and efficiency, the \proglang{C++}, \proglang{Java},
and \proglang{Python} bindings to
Protocol Buffers are used with a compiler that translates a Protocol
Buffer schema description file (ending in \texttt{.proto}) into
language-specific classes that can be used to create, read, write and
manipulate Protocol Buffer messages.  The \proglang{R} interface, in contrast,
uses a reflection-based API that is particularly well-suited for
interactive data analysis.  
All messages in \proglang{R} have a single class
structure, but different accessor methods are created at runtime based
on the named fields of the specified message type, as described in the
next section.

@@

This section describes how to use the \proglang{R} API to create and manipulate
protocol buffer messages in \proglang{R}, and how to read and write the
binary representation of the message (often called the \emph{payload}) to files and arbitrary binary
R connections.
The two fundamental building blocks of Protocol Buffers are \emph{Messages}
and \emph{Descriptors}.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.

@@

\subsection[Importing Message Descriptors from .proto files]{Importing Message Descriptors from \texttt{.proto} files}

@@

To create or parse a Protocol Buffer Message, one must first read in 
the message type specification from a \texttt{.proto} file. The 
\texttt{.proto} files are imported using the \code{readProtoFiles}
function, which can either import a single file, all files in a directory,
or every \texttt{.proto} file provided by a particular \proglang{R} package.

After importing proto files, the corresponding message descriptors are
available from the \texttt{RProtoBuf:DescriptorPool} environment in 
the \proglang{R} search path.  This environment is implemented with the user
defined tables framework from the \pkg{RObjectTables} package
available from the OmegaHat project \citep{RObjectTables}.  Instead of
being associated with a static hash table, this environment
dynamically queries the in-memory database of loaded descriptors
during normal variable lookup.

@@

% The old section 7 in entirety:
%The \texttt{RProtoBuf} package uses the user defined tables framework
%that is defined as part of the \texttt{RObjectTables} package available
%from the OmegaHat project \citep{RObjectTables}.
%
%The feature allows \texttt{RProtoBuf} to install the
%special environment \emph{RProtoBuf:DescriptorPool} in the \proglang{R} search path.
%The environment is special in that, instead of being associated with a
%static hash table, it is dynamically queried by \proglang{R} as part of R's usual
%variable lookup. In other words, it means that when the \proglang{R} interpreter
%looks for a binding to a symbol (foo) in its search path,
%it asks to our package if it knows the binding "foo", this is then
%implemented by the \texttt{RProtoBuf} package by calling an internal
%method of the \texttt{protobuf} C++ library.

%\subsection{Importing proto files}
%In contrast to the other languages (Java, C++, Python) that are officially
%supported by Google, the implementation used by the \texttt{RProtoBuf}
%package does not rely on the \texttt{protoc} compiler (with the exception of
%the two functions discussed in the previous section). This means that no
%initial step of statically compiling the proto file into C++ code that is
%then accessed by \proglang{R} code is necessary. Instead, \texttt{proto} files are
%parsed and processed \textsl{at runtime} by the protobuf C++ library---which
%is much more appropriate for a dynamic language.

@@

However, as opposed to \proglang{R} lists, no partial matching is performed
and the name must be given entirely.
The \verb|[[| operator can also be used to query and set fields
of a messages, supplying either their name or their tag number :

@@

Protocol Buffers include a 64-bit integer type, but \proglang{R} lacks native
64-bit integer support.  A workaround is available and described in
Section~\ref{sec:int64} for working with large integer values.

@@

However, the main focus of Protocol Buffer messages is
efficiency. Therefore, messages are transported as a sequence
of bytes. The \texttt{serialize} method is implemented for
Protocol Buffer messages to serialize a message into a sequence of
bytes that represents the message.
%(raw vector in \proglang{R} speech) that represents the message.

@@

The \texttt{input} argument of \texttt{read} can also be a binary
readable \proglang{R} connection, such as a binary file connection:

@@

The \CRANpkg{RProtoBuf} package uses the S4 system to store
information about descriptors and messages.  Using the S4 system
allows the package to dispatch methods that are not
generic in the S3 sense, such as \texttt{new} and
\texttt{serialize}.
Table~\ref{class-summary-table} lists the six
primary Message and Descriptor classes in RProtoBuf.  Each \proglang{R} object
contains an external pointer to an object managed by the
\texttt{protobuf} \proglang{C++} library, and the \proglang{R} objects make calls into more
than 100 \proglang{C++} functions that provide the
glue code between the \proglang{R} language classes and the underlying \proglang{C++}
classes.

@@

The \CRANpkg{Rcpp} package
\citep{eddelbuettel2011rcpp,eddelbuettel2013seamless} is used to 
facilitate this integration of the \proglang{R} and \proglang{C++} code for these objects.
Each method is wrapped individually which alllows us to add user
friendly custom error handling, type coercion, and performance
improvements at the cost of a more verbose implementation.
The \pkg{RProtoBuf} package in many ways motivated
the development of \pkg{Rcpp} Modules \citep{eddelbuettel2013exposing},
which provide a more concise way of wrapping \proglang{C++} functions and classes
in a single entity.

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Slot} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{pointer} & External pointer to the \texttt{Message} object of the \proglang{C++} protobuf library. Documentation for the
\texttt{Message} class is available from the Protocol Buffer project page. \\
%(\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.message.html#Message}) \\
\texttt{type} & Fully qualified name of the message. For example a \texttt{Person} message
has its \texttt{type} slot set to \texttt{tutorial.Person} \\[.3cm]
\textbf{Method} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{has} & Indicates if a message has a given field.   \\
\texttt{clone} & Creates a clone of the message \\
\texttt{isInitialized} & Indicates if a message has all its required fields set\\
\texttt{serialize} & serialize a message to a file, binary connection, or raw vector\\
\texttt{clear} & Clear one or several fields of a message, or the entire message\\
\texttt{size} & The number of elements in a message field\\
\texttt{bytesize} & The number of bytes the message would take once serialized\\[3mm]
%
\texttt{swap} & swap elements of a repeated field of a message\\
\texttt{set} & set elements of a repeated field\\
\texttt{fetch} & fetch elements of a repeated field\\
\texttt{setExtension} & set an extension of a message\\
\texttt{getExtension} & get the value of an extension of a message\\
\texttt{add} & add elements to a repeated field \\[3mm]
%
\texttt{str} & the \proglang{R} structure of the message\\
\texttt{as.character} & character representation of a message\\
\texttt{toString} & character representation of a message (same as \texttt{as.character}) \\
\texttt{as.list} & converts message to a named \proglang{R} list\\
\texttt{update} & updates several fields of a message at once\\
\texttt{descriptor} & get the descriptor of the message type of this message\\
\texttt{fileDescriptor} & get the file descriptor of this message's descriptor\\
\hline
\end{tabular}
\end{small}
\caption{\label{Message-methods-table}Description of slots and methods for the \texttt{Message} S4 class}
\end{table}

@@

Descriptors describe the type of a Message.  This includes what fields
a message contains and what the types of those fields are.  Message
descriptors are represented in \proglang{R} with the \emph{Descriptor} S4
class. The class contains the slots \texttt{pointer} and
\texttt{type}.  Similarly to messages, the \verb|$| operator can be
used to retrieve descriptors that are contained in the descriptor, or
invoke pseudo-methods.

@@

\begin{table}[tbp]
\centering
\begin{small}
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Slot} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{pointer} & External pointer to the \texttt{Descriptor} object of the \proglang{C++} proto library. Documentation for the
\texttt{Descriptor} class is available from the Protocol Buffer project page.\\
%\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\texttt{type} & Fully qualified path of the message type. \\[.3cm]
%
\textbf{Method} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{as.list} & return a named
list of the field, enum, and nested descriptors included in this descriptor.\\
\texttt{asMessage} & return DescriptorProto message. \\
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested 
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\bottomrule
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of slots and methods for the \texttt{Descriptor} S4 class}
\end{table}

@@

The class \emph{FieldDescriptor} represents field
descriptors in R. This is a wrapper S4 class around the
\texttt{google::protobuf::FieldDescriptor} \proglang{C++} class.
Table~\ref{fielddescriptor-methods-table} describes the methods
defined for the \texttt{FieldDescriptor} class.

\begin{table}[tbp]
\centering
\begin{small}
\begin{tabular}{lp{10cm}}
\toprule
\textbf{Slot} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{pointer} & External pointer to the \texttt{FieldDescriptor} \proglang{C++} variable \\
\texttt{name} & Simple name of the field \\
\texttt{full\_name} & Fully qualified name of the field \\
\texttt{type} & Name of the message type where the field is declared \\[.3cm]
%
\textbf{Method} & \textbf{Description} \\
\cmidrule(r){2-2}
\texttt{as.character} & Character representation of a descriptor\\
\texttt{toString} & Character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & Return FieldDescriptorProto message. \\
\texttt{name} & Return the name of the field descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{is\_extension} & Return TRUE if this field is an extension.\\
\texttt{number} & Gets the declared tag number of the field.\\
\texttt{type} & Gets the type of the field.\\
\texttt{cpp\_type} & Gets the \proglang{C++} type of the field.\\
\texttt{label} & Gets the label of a field (optional, required, or repeated).\\
\texttt{is\_repeated} & Return TRUE if this field is repeated.\\
\texttt{is\_required} & Return TRUE if this field is required.\\
\texttt{is\_optional} & Return TRUE if this field is optional.\\
\texttt{has\_default\_value} & Return TRUE if this field has a default value.\\
\texttt{default\_value} & Return the default value.\\
\texttt{message\_type} & Return the message type if this is a message type field.\\
\texttt{enum\_type} & Return the enum type if this is an enum type field.\\
\bottomrule
\end{tabular}
\end{small}
\caption{\label{fielddescriptor-methods-table}Description of slots and
  methods for the \texttt{FieldDescriptor} S4 class}
\end{table}

@@

If we try to set an int64 field in \proglang{R} to double values, we lose
precision:

@@

The protocol buffer can then be read into \proglang{R} and converted to a native
R histogram object for plotting:
